---
title: "Practical 3 Solutions"
subtitle: "Session 3: One way ANOVA"
author: "SB Galvin"
output:
  html_document:
    df_print: paged
---

Read in the `Practical3.csv` data file and perform a One way ANOVA to determine if Condition membership influences Wisdom

Conditions: 

- Immersed Self   SI
- Immersed Other  OI
- Distant Self    SO
- Distant Other   OD


##1 Read in data
```{r}
data  <-  read.csv("Practical3.csv")
```

#### View the data
```{r}
dfp3 <- data.frame(data$CONDITION, data$WISDOM)
names(dfp3)  <- c("CONDITION", "WISDOM")
dfp3
```

##2 Summarise Data
```{r}
library(summarytools)
by(dfp3,                                      # Tell R where the data is
   dfp3$CONDITION,                            # Tell R what/where the category is
   descr)
```

##3 Plot Data (Boxplot)
```{r}
boxplot(WISDOM ~ CONDITION, data = dfp3)
```

Histogram
```{r}
hist(dfp3$WISDOM)
```


##4 Perform One Way ANOVA model fit
```{r}
fit.av  <- aov(WISDOM ~ CONDITION, data = dfp3)
```


##5 Summarise Model
```{r}
summary(fit.av)
```

##6 Plot Diagnostic plots
```{r}
plot(fit.av)
```

##7 Plot residuals and test normality

#### Create Residuals
```{r}
fit.res  <-  residuals(fit.av)
```

#### Plot Residuals
```{r}
hist(fit.res)
```

#### Test Residual Normality
```{r}
shapiro.test(fit.res)
```


##8 Perform post-hoc comparisons
```{r}
TukeyHSD(fit.av)
```

