---
title: "Introduction to R and Rstudio"
subtitle: "Session3: One way ANOVA"
author: "SB Galvin"

output: html_notebook
---

##0 Data Preparation

First set up some options

```{r}
options(scipen=999)
set.seed(42) # This should make the data reproducable
```

Now generate some simulated data for illustrative use
```{r}
A  <-  rnorm(100, 100, 15)
B  <-  rnorm(100, 75, 15)
C  <-  rnorm(100, 115, 14)
D  <-  rnorm(100, 121, 18)
E  <-  rnorm(100, 112, 15)

df.wide  <-  data.frame(A, B, C, D, E)
df.wide
```
Now we will use the `reshape2` package to transform the wide data frame into a long data frame
```{r}
if(!require(reshape2)) {
	install.packages("reshape2")       # Function for installing a package if it's not already installed
}
```

The `melt()` function transforms data from wide to long. 
Important to note that for wide data with multiple variables per person you will need an argument for an `id` variable.
In this case we have two variables, group and response, and the row numbers in df.wide are meaningless
```{r}
df.long  <- melt(df.wide, 
                 variable.name = "Group", 
                 value.name = "Response")
df.long
```

##1 Summarising and Plotting Raw Data

### Summaries
```{r}
if(!require(summarytools)) {
	install.packages("summarytools")    # Function for installing a package if it's not already installed
}

```

Using the `desc()` function
```{r}
by(df.long,                                      # Tell R where the data is
   df.long$Group,                            # Tell R what/where the category is
   descr)
```


Visualisation is a useful tool to present qualitative features of data.

### Histogram of Response variable
```{r}
hist(df.long$Response,
     main = "Histogram: Response Variable",            # Main title
     xlab="Response Variable",                         # X label
     ylab = "Frequency")                               # Y label
abline(v= mean(df.long$Response),                      # Vertical Line
       col='tomato',                                
       lty='dashed',
       lwd=3)
```

We can also plot the distribtuion against a hypothetical normal curve by generating data (xfit) and normalised values (yfit), and then plotting using the `lines()` function
```{r}
#pdf("S3responsehist.pdf")
hist(df.long$Response,
     main = "Density: Response Variable",              # Main title
     xlab="Response Variable",                         # X label
     ylab = "Density",                                 # Y label
     freq=FALSE)                                       # Turn off Frequency labels for Y axis
abline(v= mean(df.long$Response),                      # Vertical Line
       col='tomato',                                
       lty='dashed',
       lwd=3)
# Add normal curve
xfit<-seq(min(df.long$Response),max(df.long$Response),length=40)
yfit<-dnorm(xfit,mean=mean(df.long$Response),sd=sd(df.long$Response))
lines(xfit, yfit, col="tomato", lwd=3, lty=8)
#dev.off()
```

### Boxplot
Boxplot of data. In this boxplot I have coloured the edges of the boxes, rather than change the fill colour. This is done using the `border` argument. If I wanted to colour the boxes, I could change the name of the argument to `col`
```{r}
#pdf("S3box.pdf")
# boxplot
boxplot(Response ~ Group,                                # Formula
        data = df.long,                                  # Data
        border = c("chocolate2",                         # Group A
                "aquamarine3",                           # Group B
                "darkmagenta",                           # Group C
                "cornflowerblue",                        # Group D
                "hotpink1"),                             # Group E  
        main="Boxplot of Response by Groups",            # Plot title
        boxlwd=3,                                        # Boxplot line weight
        xlab="Groups",                                   # X label
        ylab = "Response")                               # Y label
# Add a line for the grand mean
abline(h=mean(df.long$Response),                         # Horozontal intercept
       col="tomato",                                     # line colour
       lwd=2,                                            # Line weight
       lty="dashed")                                     # Linetype
#dev.off()
```



### GGplot 2 Boxplot with marginal density plot
Load ggplot2
```{r}
if(!require(ggplot2)|!require(ggExtra)) {
	install.packages(c("ggplot2", "ggExtra"))    # Function for installing a package if it's not already installed
}
```


Now draw in your data
```{r}
ggplot(df.long)+
  geom_boxplot(aes(x=Group, y= Response),
               colour=c("chocolate2",                            # Group A
                        "aquamarine3",                           # Group B
                        "darkmagenta",                           # Group C
                        "cornflowerblue",                        # Group D
                        "hotpink1"),                             # Group E
               size = 1.5)+                            
  geom_hline(yintercept = mean(df.long$Response), 
             colour="tomato", 
             linetype='dashed',
             size = 2) +
  theme_classic()                                                # Plain ggplot2 theme
```


```{r}
##Box plot backing
ggplot(df.long,aes(Group,Response, fill=Group))+
  geom_violin()+
  scale_fill_manual(values = c("chocolate2",                            # Group A
                               "aquamarine3",                           # Group B
                               "darkmagenta",                           # Group C
                               "cornflowerblue",                        # Group D
                               "hotpink1")) +                           # Group E
  stat_summary(fun.y=median, geom="point", size=3, color="white")+
  geom_boxplot(width=.2, fill='white', colour="grey10")+
    geom_hline(yintercept = mean(df.long$Response), 
             colour="grey35", 
             linetype='dashed',
             size = 1.5)+
  theme_classic()+
  theme(legend.position="none") ->> pVi2
pVi2
```

###Normal distribution of Response data
```{r}
 ggplot()+
  geom_line(aes(x=xfit, y = yfit),                   # Input x and y data
            colour="grey35",                         # line colour
            size = 1.5)+
  geom_vline(xintercept = mean(df.long$Response),    # x axis intercept
             size = 1.5,
             colour= "grey35",                       # Colour
             linetype = "dashed")+
  coord_flip()+                                      # Rotate the plot
  theme_void() ->> pNO                               # Add a gg theme ->> save as plot object

pNO

```

###Marginal Plots
We can plot the response data distribution against the violin plot.
```{r}
library(cowplot)
#pdf("ViolinPlot2.pdf")
plot_grid(pVi2, pNO, ncol = 2, rel_widths = c(3.3, .7), align = "h")
#dev.off()
```


##2 Assumptions

###Homogeneity of variance
Base R has a number of tests for Homogeneity of variance

The Bartlett test
```{r}
# Bartlett Test of Homogeneity of Variances
bartlett.test(Response ~ Group, 
              data = df.long)
```

The Figner-Lillen test can be used for non-parametric data
```{r}
# Figner-Killeen Test of Homogeneity of Variances
fligner.test(Response ~ Group, 
              data = df.long)
```


###Normality
We can visualise data using a Quantile-Quantile plot
```{r}
qqnorm(df.long$Response)
qqline(df.long$Response)
```


##3 ANOVA Model

ANOVA residuals don't have to be anywhere close to normal in order to fit the model. However, near-normality of the residuals is essential for p-values computed from the F-distribution to be meaningful.

### ANOVA
```{r}
model.av  <-  aov(Response ~ Group, data = df.long)
```

### Summary
```{r}
summary(model.av)
```

#### Post-hoc tests
Tukey HSD
```{r}
TukeyHSD(model.av)
```

#### Pairwise t-tests with Bonferonni corrections
Must use the `$` subset method
```{r}
pairwise.t.test(df.long$Response, df.long$Group, 
                p.adj="bonferroni", paired=FALSE)
```

We can also compute 95% confidence intervals
```{r}
confint(model.av)
```

#### Diagnostic Plots
R produces diagnostic plots with ease
```{r}
plot(model.av)
```

### Effect Size
Eta-squared, Partial Eta and Omega-squared using the `sj-stats` package
```{r}
if(!require(sjstats)) {
	install.packages("sjstats")    # Function for installing a package if it's not already installed
}
```
#### Eta - Squared
```{r}
library(sjstats)
eta_sq(model.av)
```
#### Partial Eta-Squared
One factor with 5 levels
```{r}
eta_sq(model.av,
       partial = TRUE)
```
#### Omega-Squared
```{r}
omega_sq(model.av)
```


### Residuals
A major assumption of the ANOVA model is that the model residuals are not systematic and are instead normally (randomly) distributed, allowing us to specify a cause for a response such that **GROUP MEMBERSHIP --> RESPONSE**

Residuals
```{r}
model.resid  <-  residuals(model.av)
```

We can now check to see if our model residuals are normally distributed
```{r}
shapiro.test(model.resid)
```

We can plot the model residuals using the hist funtion
```{r}
#pdf("S3ModelResid.pdf")
hist(model.resid,
     col="grey88",
     main = "Histogram: Model Residuals and Normal Density Curve",
     freq = FALSE)
abline(v=mean(model.resid), col="blue", lty=8, lwd=3)
xfit<-seq(min(model.resid),max(model.resid),length=40)
yfit<-dnorm(xfit,mean=mean(model.resid),sd=sd(model.resid))
lines(xfit, yfit, col="blue", lwd=3, lty=8)
#dev.off()
```

GGplot2 scatterplot  to explain the residuals
```{r}
df.long2  <- df.long
df.long2$Resid  <-  df.long2$Response - mean(df.long2$Response)

ggplot(df.long2, aes(fill=Group))+
  geom_jitter(aes(Group, Response), shape = 21)+
  geom_segment(data=df.long2, aes(x=Group, xend=Group, y=Response, yend= Response+Resid))+
  theme_classic() -> jitplt

#pdf("S3jit.pdf")
jitplt
#dev.off()
```


### ANOVA table
The `anova()` function produces an ANOVA table you can use for a report (this can be written as an csv file)
```{r}
anova(model.av)
```

